Challenges to transboundary fisheries management in North America under climate change

### 2020-12-14 Ecol/Manuscript_Draft.Rmd
standarized stocks and species

where *y* is year, *r* is region, *s* is grid, *n* is total number of grids in  the region, and $\hat{MCP}$ is the MCP of each species. In the case of the GoMa, region was defined as the $0.5^{\circ}$ x $0.5^{\circ}$ grid-cell within the specific NAFO divisional zones. For the IPHC analysis, region was defined as the Commission's regulatory areas (Figure \@ref(fig:ocFigure1)). I then averaged the values in three time periods (*t*) to reduce temporal model sensitivity. Thus, I computed the regional percentage change in MCP ($\Delta MCP$) as follows:

@@

In addition, I borrowed the concept of “threat point” from game theory defined as the minimum payoff that a player is willing to receive in order to cooperate with other players [@Sumaila:2020tp]. Thus, I estimated the change in the ($\Delta MCP$) (threat point) that each country (players) would have for each species (hereafter referred as stock-share ratio), for both the IPHC and the GoMa. The stock-share ratio can be seen as the proportion of the stock’s distribution within the study area that each country has. For this, I first modified equation \@ref(eq:ocone), to estimate the aggregated yearly mean MCP of each species per region. I then averaged the results by the same previously motioned periods (present, mid and end of the $21^{st}$ century). Next, for each species I estimated the stock-share ratio  ($\alpha_s$) that each region had during each time period:

@@

Where $\theta_{rt}$* is the species' aggregated $\hat{MCP}$ of each region at time period *t*, and $\delta_{ts}$ is the species' aggregated $\hat{MCP}$ of the whole specie's distribution within the study area at the same time period. Finally, I estimated the percentage change in stock-share ratio substituting $X_{t_0}$ and $X_{t}$ by $\alpha_{t0}$ and $\alpha_t$, respectively in equation \@ref(eq:octwo). The process was carried out for each ESM and results presented as average $\pm$ standard deviation (sd). All of the analysis was done in the statistical software R version 3.5.2 (2018-12-20) with the associated packages, data.table [@Packagedatatable:2019uh], ggrepel [@PackageggrepelAut:2018to], gridExtra [@PackagegridExtraM:2017wx], knirt [@PackageknitrAGen:2020wt], RColorBrewer [@PackagerColorBrewer:2014uw], sf [@PackagesfSimpleF:2018vp], and tidyverse [@PackagetidyverseE:2017vq]. All code is available at http://www.github.com/jepa/OC_Transboundary.

@@

### Projected change to species managed by the IPHC

@@

```{r ocFigure2, eval = T, echo = F, fig.cap= "Percentage change of MCP for species managed by the IPHC for mid 21st century (2041-2060) relative to present 2005-2014 under a A) high emission scenario and B) low emission scenario. Labels marked with * represent regions where models do not agree in direction of change."}

@@

### Projected change to species managed under the Gulf of Maine arrangement

While some regulatory areas of the IPHC will see an incremental increase in Pacific halibut MCP, the results for the Gulf of Maine show an overall decrease in MCP by 2050, regardless of the climate change scenario or ESM (Figure \@ref(fig:ocFigure4)), intensifying by the end of the $21^{st}$ century (Figure \@ref(fig:ocFigureS12)). For cod and haddock, MCP will decrease within the whole Gulf with no apparent win for any country in reference to the current period (Figure \@ref(fig:ocFigure4)). For yellowtail flounder, despite an overall reduction, some discrete areas are expected to increase with no particular pattern and high uncertainty, as ESMs in these regions do not agree in the direction of change (Figure \@ref(fig:ocFigureS23)). Despite the overall reduction in MCP for all three species in comparison to current values, there is a benefit of achieving a low emission scenario as reductions intensify under the high emission scenario.

@@

Despite the expected decrease in MCP for the region, changes in the stock-share ratio of species within the Gulf of Maine show different outcomes dependent on the climate change scenario and species in question. Following a high emission path will negatively affect mostly Canada’s share of yellowtail flounder and in less degree haddock, with an increase of cod share. Under the low emission scenario, haddock and cod share patterns intensify, while yellowtail flounder’s share approaches almost no change (Figure \@ref(fig:ocFigure5)). Such pattern is likely due to the combination of the bathymetry or the Gulf, the warming gradient, and the species distribution (see Discussion).

@@

The results of the present study suggest that climate change will alter the MCP of jointly managed transboundary fish stocks in North America consequently altering Canada’s and the US’s species’ stock-share ratio, regardless of the climate change scenario. These results are aligned with regional [@Morley:2018fn] projections suggesting that climate change will push marine species towards the poles and deeper water [@Pinsky:2013ab] in search of their ecological niche [@Poloczanska:2016kk]. Moreover, IPHC data[^IPHC] suggest that some of these shifts are already happening. For example, since 2010, the distribution proportion of Pacific halibut has increased from 9% to 11% in region 2B, from 7.5% to 13% in region 2C, and from 12.3% to 13.5% in region 4CDE. On the other hand, regions 3A and 3B have seen the largest decreases in the IPHC regulatory areas since 2010, from 35.3% to 30.6% and 20.6% to 15.9%, respectively. Similarly, in the Gulf of Maine, the projected stock-share gain of yellowtail flounder and haddock by the US (Figure \@ref(fig:ocFigure5)) follows a historical trend where in 2019, Canada's stock-share decreased from 35% to 32% and 60% to 40% relative to 2010, respectively [@Trinko:2019ou].

@@

Geographic barriers [@Cheung:2015fm; @Kleisner:2016je], local temperature gradients [@Pinsky:2013ab], species interactions and human activities [@Serpetti:2017he] might change the rate and direction of species shifts. For the IPHC, geographic barriers might induce a westward increase of stock-share in IPHC regions where species can only migrate northward into the Arctic Ocean through the Bering Sea and Bering Strait [@Cheung:2015fm] (Figure \@ref(fig:ocFigure3)). In the Gulf of Maine, future projections could be a response to a temperature gradient shift combined with geographic barriers as southern waters are deeper and warming slower than northern waters according to the ESMs (Figure \@ref(fig:ocFigureS22)). Moreover, Maine has seen its landings of yellowtail flounder increase at the expenses of southern states [@Pinsky:2012kq]. This could be influencing the US gain in MCP in the GoMa in relation to Canada as species shift their distribution from lower latitudes naturally reaching the US (lower) region first. As the effects of climate change endure, even with high mitigation, joint plans should prepare to face changes in the stock-share ratio of transboundary stocks along both coast of North America.

@@

Side payments have been previously used to address changes in species distribution, including cases caused by environmental forcing. In game theory, a side payment is received by a player as a compensation from the other player in a shared resource agreement, with the premise that cooperation will result in a better overall outcome [@Bjorndal:2012vd; @Sumaila:2013hv]. Side payments do not have to be in monetary form and are widely used in transboundary stocks around the world. For example, Norway and Russia have implemented a quota swap strategy for jointly managed stocks of cod, haddock and capelin (*Mallotus villosus*) in the Barents Sea [@FAO:2020vw]. Similarly, species' quota swaps are allowed, up to a degree, within regulatory areas of the European union [@Baudron:2020dj]. Specifically, for northern European spring spawning herring (*Clupea harengus*), Norway, Iceland, Faroe Islands, Russia and the EU reached an agreement to manage the stock after its collapsed, partially due to climate variations [@Miller:2013iv]. Among the implemented rules, the agreement established a dynamic quota allocation, allowing members to fish part of their quota within Norway's EEZ, and land the catch in Norwegian ports. In North America, Canada and the US have previous history with the utilization of side payments when in the 70's Pacific salmon shifted its distribution resulting in large interceptions of Canada's salmon by Alaskan fisheries [@Miller:2013iv; @Song:2017va]. The conflict was resolved by the implementation of a conservation fund that worked as a side payment for both Canada and the state of Alaska [@Miller:2013iv; @Song:2017va]. The potential adaptation of side payments in terms of quota swaps or allocating EEZ-fishing rights across the Gulf of Maine EEZs could be a potential solution as stocks shift due to climate change.

@@

Models are attempts to represent reality (in our case a future reality) based on observational data, previously established theory, and future scenarios, and are thus, subjected to different degrees of uncertainty [@Payne:2016gn]. An ensemble of models is a way to present a more robust result that accounts for differences in the structural composition of each model [@Cheung:2016gv]. In here, I used three ESMs to project future changes in species maximum catch potential. The levels of uncertainty related to the ESMs differ among case studies. Overall, results for the Gulf of Maine agree with a reduction in MCP of all three species. However, some discrete areas show a positive change for yellowtail flounder by mid $21^{st}$ century (Figure \@ref(fig:ocFigure3)), mainly driven by the GFDL model (Figure \@ref(fig:ocFigureS23)). Potential model artifacts could also be contributing to the results, especially in the northern part of the study area (Bay of Fundy) as most disagreeing grids are covered by land, which could be influencing the results. In contrast, considerable uncertainty exists in the change of MCP along the IPHC Convention area shown by a disagreement between ESMs (Figure \@ref(fig:ocFigureS24)). Off the coast of British Columbia, increasing temperature trends are consistent among ESMs, however, other processes such as acidification and deoxygenation are still not well understood from British Columbia to the Gulf of Alaska  [@TalloniAlvarez:2019gz]. Moreover, considerable uncertainty exists along the Bering Sea [@Douglas:2010Mb] and Antarctic Pacific regarding the extent and intensity of future sea-ice reduction under climate change [@Steiner:2015hc; @IPCC:2019tn]. Regarding the DBEM, its structural uncertainty has been previously tested for agreement against commonly used species distribution algorithms such as Maxent [@Phillips:2006ff] and AquaMaps [@Kaschner:2011hi; @Ready:2010wr] resulting in no qualitative differences in trends between algorithms [@Cheung:2016jd]. Finally, is worth mentioning that future changes to species distributions could be influenced by factors not captured by our model such as interactions between species [@Pecl:2017hu], adaptation of species to environmental changes, and anthropogenic factors [@Serpetti:2017he]. However, these factors are expected to increase the rate of range-shifting of the species making our results conservative [@Cheung:2010dt; @Serpetti:2017he].

@@

```{r FigureS11, eval = F, echo = F, fig.cap = "Percentage change of MCP for species managed by the IPHC for end-century (2081–2100) relative to 2005–2014 under a A) high emission scenario and B) low emission scenario. Labels marked with * represent regions where models do not agree in direction of change."}

>>>>

where *y* is year, *r* is region, *s* is grid, *n* is total number of grids in  the region, and $\hat{MCP}$ is the MCP of each stock. In the case of the GoMa, region was defined as the $0.5^{\circ}$ x $0.5^{\circ}$ grid-cell within the specific NAFO divisional zones. For the IPHC analysis, region was defined as the Commission's regulatory areas (Figure \@ref(fig:ocFigure1)). I then averaged the values in three time periods (*t*) to reduce temporal model sensitivity. Thus, I computed the regional percentage change in MCP ($\Delta MCP$) as follows:

@@

In addition, I borrowed the concept of “threat point” from game theory defined as the minimum payoff that a player is willing to receive in order to cooperate with other players [@Sumaila:2020tp]. Thus, I estimated the change in the ($\Delta MCP$) (threat point) that each country (players) would have for each stock (hereafter referred as stock-share ratio), for both the IPHC and the GoMa. The stock-share ratio can be seen as the proportion of the stock’s distribution within the study area that each country has. For this, I first modified equation \@ref(eq:ocone), to estimate the aggregated yearly mean MCP of each stock per region. I then averaged the results by the same previously motioned periods (present, mid and end of the $21^{st}$ century). Next, for each stock I estimated the stock-share ratio ($\alpha_s$) that each region had during each time period:

@@

Where $\theta_{rt}$* is the stock's aggregated $\hat{MCP}$ of each region at time period *t*, and $\delta_{ts}$ is the stock's aggregated $\hat{MCP}$ of the whole stock's distribution within the study area at the same time period. Finally, I estimated the percentage change in stock-share ratio substituting $X_{t_0}$ and $X_{t}$ by $\alpha_{t0}$ and $\alpha_t$, respectively in equation \@ref(eq:octwo). The process was carried out for each ESM and results presented as average $\pm$ standard deviation (sd). All of the analysis was done in the statistical software R version 3.5.2 (2018-12-20) with the associated packages, data.table [@Packagedatatable:2019uh], ggrepel [@PackageggrepelAut:2018to], gridExtra [@PackagegridExtraM:2017wx], knirt [@PackageknitrAGen:2020wt], RColorBrewer [@PackagerColorBrewer:2014uw], sf [@PackagesfSimpleF:2018vp], and tidyverse [@PackagetidyverseE:2017vq]. All code is available at http://www.github.com/jepa/OC_Transboundary.

@@

### Projected change to stocks managed by the IPHC

@@

```{r ocFigure2, eval = T, echo = F, fig.cap= "Percentage change of MCP for stocks managed by the IPHC for mid 21st century (2041-2060) relative to present 2005-2014 under a A) high emission scenario and B) low emission scenario. Labels marked with * represent regions where models do not agree in direction of change."}

@@

### Projected change to stocks managed under the Gulf of Maine arrangement

While some regulatory areas of the IPHC will see an incremental increase in Pacific halibut MCP, the results for the Gulf of Maine show an overall decrease in MCP by 2050, regardless of the climate change scenario or ESM (Figure \@ref(fig:ocFigure4)), intensifying by the end of the $21^{st}$ century (Figure \@ref(fig:ocFigureS12)). For cod and haddock, MCP will decrease within the whole Gulf with no apparent win for any country in reference to the current period (Figure \@ref(fig:ocFigure4)). For yellowtail flounder, despite an overall reduction, some discrete areas are expected to increase with no particular pattern and high uncertainty, as ESMs in these regions do not agree in the direction of change (Figure \@ref(fig:ocFigureS23)). Despite the overall reduction in MCP for all three stocks in comparison to current values, there is a benefit of achieving a low emission scenario as reductions intensify under the high emission scenario.

@@

Despite the expected decrease in MCP for the region, changes in the stock-share ratio of stocks within the Gulf of Maine show different outcomes dependent on the climate change scenario and stock in question. Following a high emission path will negatively affect mostly Canada’s share of yellowtail flounder and in less degree haddock, with an increase of cod share. Under the low emission scenario, haddock and cod share patterns intensify, while yellowtail flounder’s share approaches almost no change (Figure \@ref(fig:ocFigure5)). Such pattern is likely due to the combination of the bathymetry or the Gulf, the warming gradient, and the stock's distribution (see Discussion).

@@

The results of the present study suggest that climate change will alter the MCP of jointly managed transboundary fish stocks in North America consequently altering Canada’s and the US’s stock’s stock-share ratio, regardless of the climate change scenario. These results are aligned with regional [@Morley:2018fn] projections suggesting that climate change will push marine species towards the poles and deeper water [@Pinsky:2013ab] in search of their ecological niche [@Poloczanska:2016kk]. Moreover, IPHC data[^IPHC] suggest that some of these shifts are already happening. For example, since 2010, the distribution proportion of Pacific halibut has increased from 9% to 11% in region 2B, from 7.5% to 13% in region 2C, and from 12.3% to 13.5% in region 4CDE. On the other hand, regions 3A and 3B have seen the largest decreases in the IPHC regulatory areas since 2010, from 35.3% to 30.6% and 20.6% to 15.9%, respectively. Similarly, in the Gulf of Maine, the projected stock-share gain of yellowtail flounder and haddock by the US (Figure \@ref(fig:ocFigure5)) follows a historical trend where in 2019, Canada's stock-share decreased from 35% to 32% and 60% to 40% relative to 2010, respectively [@Trinko:2019ou].

@@

Geographic barriers [@Cheung:2015fm; @Kleisner:2016je], local temperature gradients [@Pinsky:2013ab], species interactions and human activities [@Serpetti:2017he] might change the rate and direction of species shifts. For the IPHC, geographic barriers might induce a westward increase of stock-share in IPHC regions where the stock can only migrate northward into the Arctic Ocean through the Bering Sea and Bering Strait [@Cheung:2015fm] (Figure \@ref(fig:ocFigure3)). In the Gulf of Maine, future projections could be a response to a temperature gradient shift combined with geographic barriers as southern waters are deeper and warming slower than northern waters according to the ESMs (Figure \@ref(fig:ocFigureS22)). Moreover, Maine has seen its landings of yellowtail flounder increase at the expenses of southern states [@Pinsky:2012kq]. This could be influencing the US gain in MCP in the GoMa in relation to Canada as stocks shift their distribution from lower latitudes naturally reaching the US (lower) region first. As the effects of climate change endure, even with high mitigation, joint plans should prepare to face changes in the stock-share ratio of transboundary stocks along both coast of North America.

@@

Side payments have been previously used to address changes in stock's distribution, including cases caused by environmental forcing. In game theory, a side payment is received by a player as a compensation from the other player in a shared resource agreement, with the premise that cooperation will result in a better overall outcome [@Bjorndal:2012vd; @Sumaila:2013hv]. Side payments do not have to be in monetary form and are widely used in transboundary stocks around the world. For example, Norway and Russia have implemented a quota swap strategy for jointly managed stocks of cod, haddock and capelin (*Mallotus villosus*) in the Barents Sea [@FAO:2020vw]. Similarly, species' quota swaps are allowed, up to a degree, within regulatory areas of the European union [@Baudron:2020dj]. Specifically, for northern European spring spawning herring (*Clupea harengus*), Norway, Iceland, Faroe Islands, Russia and the EU reached an agreement to manage the stock after its collapsed, partially due to climate variations [@Miller:2013iv]. Among the implemented rules, the agreement established a dynamic quota allocation, allowing members to fish part of their quota within Norway's EEZ, and land the catch in Norwegian ports. In North America, Canada and the US have previous history with the utilization of side payments when in the 70's Pacific salmon shifted its distribution resulting in large interceptions of Canada's salmon by Alaskan fisheries [@Miller:2013iv; @Song:2017va]. The conflict was resolved by the implementation of a conservation fund that worked as a side payment for both Canada and the state of Alaska [@Miller:2013iv; @Song:2017va]. The potential adaptation of side payments in terms of quota swaps or allocating EEZ-fishing rights across the Gulf of Maine EEZs could be a potential solution as stocks shift due to climate change.

@@

Models are attempts to represent reality (in our case a future reality) based on observational data, previously established theory, and future scenarios, and are thus, subjected to different degrees of uncertainty [@Payne:2016gn]. An ensemble of models is a way to present a more robust result that accounts for differences in the structural composition of each model [@Cheung:2016gv]. In here, I used three ESMs to project future changes in species maximum catch potential. The levels of uncertainty related to the ESMs differ among case studies. Overall, results for the Gulf of Maine agree with a reduction in MCP of all three stocks. However, some discrete areas show a positive change for yellowtail flounder by mid $21^{st}$ century (Figure \@ref(fig:ocFigure3)), mainly driven by the GFDL model (Figure \@ref(fig:ocFigureS23)). Potential model artifacts could also be contributing to the results, especially in the northern part of the study area (Bay of Fundy) as most disagreeing grids are covered by land, which could be influencing the results. In contrast, considerable uncertainty exists in the change of MCP along the IPHC Convention area shown by a disagreement between ESMs (Figure \@ref(fig:ocFigureS24)). Off the coast of British Columbia, increasing temperature trends are consistent among ESMs, however, other processes such as acidification and deoxygenation are still not well understood from British Columbia to the Gulf of Alaska  [@TalloniAlvarez:2019gz]. Moreover, considerable uncertainty exists along the Bering Sea [@Douglas:2010Mb] and Antarctic Pacific regarding the extent and intensity of future sea-ice reduction under climate change [@Steiner:2015hc; @IPCC:2019tn]. Regarding the DBEM, its structural uncertainty has been previously tested for agreement against commonly used species distribution algorithms such as Maxent [@Phillips:2006ff] and AquaMaps [@Kaschner:2011hi; @Ready:2010wr] resulting in no qualitative differences in trends between algorithms [@Cheung:2016jd]. Finally, is worth mentioning that future changes to species distributions could be influenced by factors not captured by our model such as interactions between species [@Pecl:2017hu], adaptation of species to environmental changes, and anthropogenic factors [@Serpetti:2017he]. However, these factors are expected to increase the rate of range-shifting of the species making our results conservative [@Cheung:2010dt; @Serpetti:2017he].

@@

```{r FigureS11, eval = F, echo = F, fig.cap = "Percentage change of MCP for stocks managed by the IPHC for end-century (2081–2100) relative to 2005–2014 under a A) high emission scenario and B) low emission scenario. Labels marked with * represent regions where models do not agree in direction of change."}
